Speed and Precision in Range Analysis

### 2012-04-11 doc/Manuscript/SBLP/paper.tex
Alterei meu e-mail
Alterando o e-mail do Igor
Added discussion about the widening operator to the paper.

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
\email{\{victorsc,igor.rafael,raphael.hernani,fernando\}@dcc.ufmg.br}
}

@@

% Define the lattice, the constraints and the valuation I.
\paragraph{The Interval Lattice.}
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the complete lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
For any $x > -\infty$ we define:

@@

\paragraph{A Holistic View of our Range Analysis Algorithm.}
We perform range analysis in a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.
In Section~\ref{sec:design} we discuss in more details these tradeoffs.
The last phase, which we call the {\em micro algorithm}, happens per
strong component; however, if we opted for not building these components,
then it happens once for the entire constraint graph.
Nevertheless, the use of strongly connected components
is so essential for performance and precision, as we show in
Section~\ref{sub:sccs}, that it is considered optional only because we
can easily build our implementation without this module.

@@

\paragraph{The Micro Algorithm.}
We find the solution given in Figure~\ref{fig:overall_view}(e) in a
process that we call the micro algorithm.
This phase is further divided into three sub-steps:
(i) growth analysis;
(ii) future resolution and
(iii) narrowing analysis.

@@

\paragraph{Time and Memory Complexity}

@@

\paragraph{Precision}

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation are responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our runtime, because it increases the
size of the target program.
Nevertheless, the slowdown is inexpressive.
We chose to run our analysis in programs in the Extended Static Single
Assignment (e-SSA) form~\cite{Bodik00}.
The alternative, which gives us a faster, albeit imprecise, analysis, is to run
our algorithm in programs in Static Single Assignment (SSA)
form~\cite{Cytron91}.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

@@

@@

>>>>

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
\email{\{victorsc,raphael.ernani,igor.rafael,fernando\}@dcc.ufmg.br}
}

@@

% Define the lattice, the constraints and the valuation I.
\noindent
\textbf{The Interval Lattice.}
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the complete lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots < -2 < -1 < 0 < 1 < 2 < \ldots +\infty$.
For any $x > -\infty$ we define:

@@

\noindent
\textbf{A Holistic View of our Range Analysis Algorithm.}
We perform range analysis in a number of steps.
First, we convert the program to a suitable intermediate representation that
makes it easier to extract constraints.
From these constraints, we build a dependence graph that allows us to do
range analysis sparsely.
Finally, we solve the constraints applying different fix-point iterators on
this dependence graph.
Figure~\ref{fig:algorithm} gives a global view of this algorithm.
Some of the steps in the algorithm are optional.
They improve the precision of the range analysis, at the expense of a longer
running time.
In Section~\ref{sec:design} we discuss in more details these tradeoffs.
The last phase, which we call the {\em micro algorithm}, happens per
strong component; however, if we opted for not building these components,
then it happens once for the entire constraint graph.
Nevertheless, the use of strongly connected components
is so essential for performance and precision, as we show in
Section~\ref{sub:sccs}, that it is considered optional only because we
can easily build our implementation without this module.

@@

\noindent
\textbf{The Micro Algorithm.}
We find the solution given in Figure~\ref{fig:overall_view}(e) in a
process that we call the micro algorithm.
This phase is further divided into three sub-steps:
(i) growth analysis;
(ii) future resolution and
(iii) narrowing analysis.

@@

\noindent
\textbf{Time and Memory Complexity}

@@

\noindent
\textbf{Precision}

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation are responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our runtime, because it increases the
size of the target program.
We have tried our analysis in two different program representations: the
Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
Single Assignment (e-SSA) form~\cite{Bodik00}.
The SSA form gives us a faster, albeit more imprecise, analysis.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

@@

We have implemented the widening operator used in the growth analysis in
two different ways.
The first way, which we call {\em simple}, is based on Cousot and
Cousot's original widening operator~\cite{Cousot77}.
This operator is shown in Figure~\ref{fig:widening}, and it is the one
used in Figure~\ref{fig:ex_partition_grow_crop}(b).
The second widening strategy, which we call {\em jump-set widening} consists
in using the constants that appear in the program text, in sorted order, as
the next limits of each interval after widening is applied.
This operator is common in implementations of range
analysis~\cite[p.228]{Nielson99}.
There are situations in which jump-set widening produces better results
than the simple operator.
Figure~\ref{fig:jumpSet} shows an example taken from the code of
SPEC CPU \texttt{bzip2}.
Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
is given in Figure~\ref{fig:jumpSet}(b).
The result of applying the simple operator is shown in
Figure~\ref{fig:jumpSet}(c).
Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
instead of the lattice in Figure~\ref{fig:widening}(Right).
This lattice would yield the result given in Figure~\ref{fig:jumpSet}(e),
which is more precise.

\begin{figure}[t!]
\begin{center}
\begin{tabular}{c@{\hspace{1.5cm}}c}
\begin{minipage}{2cm}
\includegraphics{images/growth_lattice}
\end{minipage}
&
\begin{minipage}{8cm}
\begin{small}
\begin{eqnarray*}
\begin{array}{c@{\hspace{0.5cm}}c}
\inferrule{I[Y] = [\bot, \bot]}{I[Y] \leftarrow e(Y)}
&
\inferrule{\lb{e(Y)} < \lb{I[Y]} \\ \ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [-\infty, +\infty ]}
\\
\\
\inferrule{\lb{e(Y)} < \lb{I[Y]}}
{I[Y] \leftarrow [-\infty, \ub{I[Y]}]}
&
\inferrule{\ub{e(Y)} > \ub{I[Y]}}
{I[Y] \leftarrow [\lb{I[Y]}, +\infty]}
\end{array}
\end{eqnarray*}
\end{small}
\end{minipage}
\end{tabular}
\end{center}
\caption{\label{fig:widening}
(Left) The lattice used in the simple widening strategy.
(Right) Cousot and Cousot's widening operator. We evaluate the rules from
left-to-right, top-to-bottom, and stop upon finding a pattern matching.
Again: given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/jumpSet}
\end{center}
\caption{\label{fig:jumpSet}
An example where jump-set widening is more precise.}
\end{figure}

Another way to improve the precision of growth analysis is to perform a few
rounds of abstract interpretation on the constraint graph, and, in case the
process does not reach a fix point, only then applying the widening operator.
Each round consists in evaluating all the constraints, and updating the
intervals that change from one evaluation to the other.
The simple widening strategy, coupled with one round of abstract interpretation
would give us precise results in the example from Figure~\ref{fig:jumpSet}.
We have experimented with 0 and 16 iterations before doing widening, and the
overall result, for the programs in the SPEC CPU 2006 suite is given in
Figure~\ref{fig:space}.
Figure~\ref{fig:wideningPrec} shows some of these results in more details
for a broader set of benchmarks.
In general jump-set widening improves the precision of our results in
non-trivial ways.
Nevertheless, the simple widening operator preceded by 16 rounds of
abstract interpretation in general is more precise than jump-set widening
without any cycle of pre-evaluation, as we see in Figure~\ref{fig:space}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/wideningPrec}
\end{center}
\caption{\label{fig:wideningPrec}
The impact of the widening strategy on precision.
Each bar gives precision in \%bitwidth reduciton.}
\end{figure}

@@

This paper presents what we believe is the most comprehensive evaluation of
range analysis in the literature.
Altogether we have experimented with 32 different configurations of our
range analysis algorithm.
Our implementation is
publicly available at \texttt{http://code.google.com/p/range-analysis/}.
This repository contains instructions about how to deploy and use our
implementation.
We provide a gallery of examples, including source codes,
CFGs and constraint graphs that we produce for meaningful programs at
\texttt{http://code.google.com/p/range-analysis/wiki/gallery}.
